Analyzing TA Comments on Student Writing in R

### 2019-02-05 draft.Rmd
Update

---
title: "Draft Description of Work in Progress"
author: "Dan Johnson"
date: "12/17/2018"
output: html_document
bibliography: library.bib
biblio-style: apalike
---
```{r, child="_setup.Rmd"}
```
#Overview
This page provides a draft overview of my progress. A short narrative outlines the  discrete blocks of work and how they are connected. Links connect to longer pages that document intake, munging, analysis, and visualization in depth.

@@

##Initial Dataset
My working dataset is an Excel table of 10,808 rows. One column has "TA comments" extracted from student lab reports in Spring 2018. Each TA comment is stored as a separate row of the table, and is a mixed alphanumeric string of one or more words, numbers, and punctuation. Other columns record unique report, student, and TA IDs; grade assigned to the report; other standardized information about the original report from which the comment was extracted; and the hand-coded subject and structure of each comment.

Data were anonymized in Excel prior to conversion to CSV format. Names, email addresses, and other confidential information was replaced with unique randomized alphanumeric strings.

@@

##Blocks of Work

*  Data Pre-Checks and Import
*  Initial Import and Post-Checks
*  Preliminary Exploration: Log-Odds Text Comparison
    +  Pearson Correlations of Word Frequencies: BROKEN SECTION
    +  Mapping Analyses Visually: RUNS BUT CANNOT COPY AND MODIFY. It depends tables created using the broken code above. So I cannot run this part either.
    +  Calculating and Plotting Log-Odds Pairs
*  Preliminary Exploration: Direct Comparison of Word Lists
*  Deeper Text Analysis Methods
    +  Naive Bayes [/docs/Naive_bayes.html](link)
    +  Random Forest (link)
    +  Latent Dirichlet Allocation (link)
    +  Etc. (link)
    +  Etc. (link)

##Data Pre-Checks
Prior to import, review data table in Excel. Check:

1.  Vocabulary terms used in coding columns matches criteria in "codebook_for_comments.rmd"
2.  Data have been de-identified and recoded using pre-assigned anonymous IDs.
3.  Data table headers are in snake_case, and match tabulated list below. The same list is in the file "code_column_names.rmd"

Column Number|Column Name|Column Number|Column Name
------------|----------------------|------------|----------------------
1|unique.record|15|data.ok
2|report.id|16|citation.ok
3|sort|17|interpretation.ok
4|report.title (empty)|18|organization.ok
5|student (Std\_nnn)|19|techflaws.ok
6|course|20|writing.ok
7|ta (TA\_nnn)|21|comments.incorporated
8|lab|22|ta.comment
9|tag|23|code.subject
10|type.TA|24|code.structure
11|grade.TA|25|code.locus
12|grading.time|26|code.scope
13|Rank|27|code.tone
14|hypothesis.ok|28|code.notes

@@

##Initial Import and Post-Check
```{r Libraries_1}
# Call in tidyverse, other required libraries.
library(tidyverse)
library(tidytext)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
```

```{r}
#This block reads in the full CSV into a starting dataframe named "base_data".
#Subsequent analyses call in subsets of "base_data".
base_data <- read_csv(file='data/coded_full_comments_dataset_Spring18anon.csv')

#If dataset imports correctly, there will be a dataframe of ~11,000 rows and 28 columns named as listed in the **Data Pre-Checks** table above.
```

Given some data points may be entered incorrectly by instructors or students, these post-import data checks help ensure data have been properly coded and columns contain valid entries only.

@@

#"Lab" should only have entries that match the limited keywords in codebook.
lab<-unique(base_data$lab)

#Look at the NUMBER of unique anonymous student IDs. Should be within 10% of (but not more than) the combined enrollment of the 3 target courses for the semester.
student<-unique(base_data$student)

@@

#Check that the list of "code.structure" terms extracted from the dataset and displayed here matches allowed terms in codebook.
structure<-unique(base_data$code.structure)
```

## Preliminary Exploration
The first question to answer before making deeper comparisons is **are the comments obviously different between sub-categories?** Before conducting detailed analyses I tried several iterations of different visual comparison strategies.

###Log-Odds Text Comparison
The log-odds ratios method outlined below finds and displays differences words that are more frequently found in TA comments belonging to one sub-category vs. another. I based this on a method for comparing word usage frequencies between two separate Twitter users [https://www.tidytextmining.com/twitter.html](that I found here.)

For simplicity I reduced my analysis to just the four most common comment subjects. This has been the most informative preliminary visualization I have found so far.

```{r}
# Isolate table rows to compare. Then reduce larger dataframe to only 2 required columns of data.
frequency_writing <- filter(base_data, code.subject=="2. Writing Quality"|code.subject=="3. Technical and Scientific"|code.subject=="4. Logic and Thinking")
frequency_writing.subcolumns <- frequency_writing %>% select(1, 22:23)

# Tokenize phrases, remove stop words listed in standard reference file.
base_data_tokenized <- frequency_writing.subcolumns %>% 
  unnest_tokens(word,ta.comment) %>% 
  anti_join(stop_words)

#At this point the dataframe should have "Unique.Record" in column 1, "code.subject" in column 2, and the unnested "word" in column 3. All columns are class = "character". Stop words listed in the "stop_words" R reference data file have been removed.

```

####Comparing Pearson Correlations of Word Frequencies
If the words used by TAs in their comments differ between the categories, there should be fairly low correlations overall .
```{r}
#Block below re-organizes the data for comparisons. This pattern sets "2. Writing Quality" as the first dataset, then all other data in remaining columns.
#Groups by code.subject, calculated proportional frequency in each group.
#Final step creates 4 columns: word, 2. Writing Quality, code.subject, proportion
#Stats need this format to be able to compare 2. WQ against other values.
base_data_tokenized_sorted2 <- base_data_tokenized %>% 
  group_by(code.subject) %>%
  count(code.subject, word, sort=TRUE) %>% 
  mutate(proportion = n / sum(n)) %>% 
  select(-n) %>% 
  spread(code.subject, proportion) %>% 
  gather(code.subject, proportion, `3. Technical and Scientific`:`4. Logic and Thinking`)

#The dataframe called "base_data_tokenized_sorted2" is the final 4-column dataframe for statistical analysis. Individual words are in column 1, "2. Writing Quality" frequency is in Column 2, "code.subject" is in column 3, and "proportion" is in column 4. Columns 1 and 3 are class = "character", 2 and 4 are numeric ranges.

#This structure is needed in order to compare word (in Column 1) frequencies for 2. Writing Quality (Column 2) as Y values against the frequencies of those same words (in Column 4) as X axes. WHICH subset of X values to use is coded in Column 3.
```

**Pearson correlation** between the frequency of words in Writing Quality versus Technical and Logic sub-categoris.

```{r}
cor.test(data = base_data_tokenized_sorted2[base_data_tokenized_sorted2$code.subject == "3. Technical and Scientific",], 
         ~ proportion + `2. Writing Quality`)
```

```{r}
cor.test(data = base_data_tokenized_sorted2[base_data_tokenized_sorted2$code.subject == "4. Logic and Thinking",], 
         ~ proportion + `2. Writing Quality`)
```

***
***
####START OF BROKEN SECTION
THERE IS AN ERROR HERE I CANNOT DEBUG! If I change "3. Technical and Scientific"  to "2. Writing Quality", the structure of the tibble breaks. It goes from the correctly sorted 4 columns to incorrectly sorted 3 columns. It is NOT A TYPO; I can literally copy/paste the names in and out and break/correct the code.

I cannot fix it using a different order for variables. I am at a loss for why it breaks.

```{}
# Isolate table rows to compare. Then reduce larger dataframe to only 2 required columns of data.
frequency_writing3 <- filter(base_data, code.subject=="3. Technical and Scientific"|code.subject=="2. Writing Quality"|code.subject=="4. Logic and Thinking")
frequency_writing3.subcolumns <- frequency_writing3 %>% select(1, 22:23)
```

```{}
# Tokenize phrases, remove stop words listed in standard reference file.
base_data_tokenized3 <- frequency_writing3.subcolumns %>% 
  unnest_tokens(word,ta.comment) %>% 
  anti_join(stop_words)
```

```{}
#Block below re-organizes the data for comparisons. This pattern sets "3. Writing Quality" as the first dataset, then all other data in remaining columns.
#Groups by code.subject, calculated proportional frequency in each group.
#Final step creates 4 columns: word, 2. Writing Quality, code.subject, proportion
#Stats need this format to be able to compare 2. WQ against other values.
base_data_tokenized_sorted3 <- base_data_tokenized3 %>% 
  group_by(code.subject) %>%
  count(code.subject, word, sort=TRUE) %>% 
  mutate(proportion = n / sum(n)) %>% 
  select(-n) %>% 
  spread(code.subject, proportion) %>% 
  gather(code.subject, proportion, `2. Writing Quality`:`4. Logic and Thinking`)
```

```{}
#Create groups by code.subject, calculates proportional frequency in each group.
#Final step creates 5 columns: code.subject, n, word, total, frequency.
base_data_tokenized_sorted3 <- base_data_tokenized3 %>% 
  group_by(code.subject) %>%
  count(code.subject, word, sort=TRUE) %>% 
  left_join(base_data_tokenized %>% 
              group_by(code.subject) %>% 
              summarise(total = n())) %>%
  mutate(freq = n/total)
```

Re-organize the data for comparisons. This pattern sets 3. Technical and Scientific as the first dataset, then all other data in remaining columns.
```{}
#Groups by code.subject, calculated proportional frequency in each group.
#Final step creates 4 columns: word, 3. Technical and Scientific, code.subject, proportion
#Stats need this format to be able to compare 3. TS against other values.
base_data_tokenized_sorted3 <- base_data_tokenized3 %>% 
  group_by(code.subject) %>%
  count(code.subject, word, sort=TRUE) %>% 
  mutate(proportion = n / sum(n)) %>% 
  select(-n) %>% 
  spread(code.subject, proportion) %>% 
  gather(code.subject, proportion, `2. Writing Quality`:`4. Logic and Thinking`)

#END OF BROKEN SECTION
```

I can run this block, but CANNOT COPY AND MODIFY THIS SECTION. It needs the tables that should be created using the broken code above. So I cannot run this part either.

```{}
#Create groups by code.subject, calculates proportional frequency in each group.
#Final step creates 5 columns: code.subject, n, word, total, frequency.
base_data_tokenized_sorted2 <- base_data_tokenized %>% 
  group_by(code.subject) %>%
  count(code.subject, word, sort=TRUE) %>% 
  left_join(base_data_tokenized %>% 
  group_by(code.subject) %>% 
  summarise(total = n())) %>%
  mutate(freq = n/total)
```

```{}
base_data_tokenized_sorted2v3 <- base_data_tokenized_sorted2 %>% 
  select(code.subject, word, freq) %>% 
  spread(code.subject, freq) %>%
  arrange(`2. Writing Quality`,`3. Technical and Scientific`)
```

```{}
ggplot(base_data_tokenized_sorted2v3, aes(`2. Writing Quality`,`3. Technical and Scientific`)) +
  geom_jitter(alpha = 0.1, size = 2.5, width = 0.25, height = 0.25) +
  geom_text(aes(label = word), check_overlap = TRUE, vjust = 1.5) +
  scale_x_log10(labels = percent_format()) +
  scale_y_log10(labels = percent_format()) +
  geom_abline(color = "red")
  #ggsave("Freqplot_3vs2.png", width = 10, height = 6)
```

```{}
#I can get part way around the problem by re-using the same data and calling a different subset for the Y axis. Want to add comparisons of 3 vs 4, but same tables cannot be created because dataframe keeps breaking when I try to change 2. WQ.

base_data_tokenized_sorted2v4 <- base_data_tokenized_sorted2 %>% 
  select(code.subject, word, freq) %>% 
  spread(code.subject, freq) %>%
  arrange(`2. Writing Quality`,`4. Logic and Thinking`)
```

```{}
ggplot(base_data_tokenized_sorted2v3, aes(`2. Writing Quality`,`4. Logic and Thinking`)) +
  geom_jitter(alpha = 0.1, size = 2.5, width = 0.25, height = 0.25) +
  geom_text(aes(label = word), check_overlap = TRUE, vjust = 1.5) +
  scale_x_log10(labels = percent_format()) +
  scale_y_log10(labels = percent_format()) +
  geom_abline(color = "red")
  #ggsave("Freqplot_4vs2.png", width = 10, height = 6)
```
END of broken / limping sections
***
***

###Calculating and Plotting Log-Odds Pairs

```{r}
word_ratios2v3 <- base_data_tokenized %>%
  count(word, code.subject) %>%
  group_by(word) %>%
  filter(sum(n) >= 10) %>%
  ungroup() %>%
  spread(code.subject, n, fill = 0) %>%
  mutate_if(is.numeric, funs((. + 1) / (sum(.) + 1))) %>%
  mutate(logratio = log(`2. Writing Quality`/`3. Technical and Scientific`)) %>%
  arrange(desc(logratio))

word_ratios3v4 <- base_data_tokenized %>%
  count(word, code.subject) %>%
  group_by(word) %>%
  filter(sum(n) >= 10) %>%
  ungroup() %>%
  spread(code.subject, n, fill = 0) %>%
  mutate_if(is.numeric, funs((. + 1) / (sum(.) + 1))) %>%
  mutate(logratio = log(`3. Technical and Scientific`/`4. Logic and Thinking`)) %>%
  arrange(desc(logratio))

word_ratios2v4 <- base_data_tokenized %>%
  count(word, code.subject) %>%
  group_by(word) %>%
  filter(sum(n) >= 10) %>%
  ungroup() %>%
  spread(code.subject, n, fill = 0) %>%
  mutate_if(is.numeric, funs((. + 1) / (sum(.) + 1))) %>%
  mutate(logratio = log(`2. Writing Quality`/`4. Logic and Thinking`)) %>%
  arrange(desc(logratio))
```

```{r}
word_ratios2v3 %>% 
  arrange(abs(logratio))

word_ratios3v4 %>% 
  arrange(abs(logratio))

word_ratios2v4 %>% 
  arrange(abs(logratio))
```

```{r}
word_ratios2v3 %>%
  group_by(logratio < 0) %>%
  top_n(20, abs(logratio)) %>%
  ungroup() %>%
  mutate(word = reorder(word, logratio)) %>%
  ggplot(aes(word, logratio, fill = logratio < 0)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  ylab("log odds ratio (Writing-top/Technical-bottom)") +
  scale_fill_discrete(name = "", labels = c("2. Writing Quality", "3. Technical and Scientific"))
  #ggsave("Logplot_2vs3.png", width = 6, height = 10)

word_ratios3v4 %>%
  group_by(logratio < 0) %>%
  top_n(20, abs(logratio)) %>%
  ungroup() %>%
  mutate(word = reorder(word, logratio)) %>%
  ggplot(aes(word, logratio, fill = logratio < 0)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  ylab("log odds ratio (Technical-top/Logic-bottom)") +
  scale_fill_discrete(name = "", labels = c("3. Technical and Scientific","4. Logic and Thinking"))
  #ggsave("Logplot_3vs4.png", width = 6, height = 10)

word_ratios2v4 %>%
  group_by(logratio < 0) %>%
  top_n(20, abs(logratio)) %>%
  ungroup() %>%
  mutate(word = reorder(word, logratio)) %>%
  ggplot(aes(word, logratio, fill = logratio < 0)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  ylab("log odds ratio (Writing-top/Logic-bottom)") +
  scale_fill_discrete(name = "", labels = c("2. Writing Quality", "4. Logic and Thinking"))
  #ggsave("Logplot_2vs4.png", width = 6, height = 10)

```

##Direct Comparison Using Only Words Appearing in All Subsets of Comments
I think this is less informative, but it does let me see the data more clearly.

Basic strategy is to take all comments from base_data, unnest tokens, filter stopwords, count and sort remaining terms, then write back to a new datatable called 'sortedwords.comments.all'.  Calculate TOTAL words and store. Then calculate fractional frequencies for each word and store using mutate.

I wrote this block so that, once CSV is imported, it does not require any of the code above. This is an early effort, so likely has some poorly developed code that could be refactored.

```{}
# Call in tidyverse, other required libraries.
library(tidyverse)
library(tidytext)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)

#Read in the full CSV into a starting dataframe named "base_data".
#Dataframe should have ~11,000 rows and 28 columns .
base_data <- read_csv(file='data/coded_full_comments_dataset_Spring18anon.csv')
```

```{r}
# Tokenize phrases, remove the stop words listed in R's standard reference file.
base_data_tokenized <- base_data %>% 
  unnest_tokens(word,ta.comment)

# This is an ALTERNATIVE METHOD for pulling out stopwords.  
sortedwords.comments.all <- base_data_tokenized %>%
  filter(!word %in% stop_words$word,
         !word %in% str_remove_all(stop_words$word, "'"),
         str_detect(word, "[a-z]")) %>%
  count(word, word, sort = TRUE) %>%
  ungroup()

#Calculate the number of words in total in the subset
total_words.all <- sortedwords.comments.all %>% 
  summarize(total = sum(n))

#Mutate table to calculate then append fractional word frequency in the text.
sortedwords.comments.all.fraction <- sortedwords.comments.all %>% as_tibble() %>% mutate(
  all.fraction = (n*100)/total_words.all$total
  )
```

```{r}
#Pull SUBSETs of tokenized data based on code.subject: 
base_data_tokenized.basic <- subset(base_data_tokenized, code.subject == "1. Basic Criteria", select = (1:28))
base_data_tokenized.writing <- subset(base_data_tokenized, code.subject == "2. Writing Quality", select = (1:28))
base_data_tokenized.technical <- subset(base_data_tokenized, code.subject == "3. Technical and Scientific", select = (1:28))
base_data_tokenized.logic <- subset(base_data_tokenized, code.subject == "4. Logic and Thinking", select = (1:28))
base_data_tokenized.praise <- subset(base_data_tokenized, code.subject == "5. Praise or Concern", select = (1:28))
base_data_tokenized.misconduct <- subset(base_data_tokenized, code.subject == "6. Misconduct", select = (1:28))
base_data_tokenized.narrative <- subset(base_data_tokenized, code.subject == "12. Narrative Comments", select = (1:28))
base_data_tokenized.sciname <- subset(base_data_tokenized, code.subject == "Scientific Name", select = (1:28))
```

For each subset, filter stopwords, count words and total, then calculate frequencies.

```{r}
sortedwords.basic <- base_data_tokenized.basic %>%
  filter(!word %in% stop_words$word,
         !word %in% str_remove_all(stop_words$word, "'"),
         str_detect(word, "[a-z]")) %>%
  count(word, word, sort = TRUE) %>%
  ungroup()

#Calculate the number of words in total in the subset
total_words.basic <- sortedwords.basic %>% 
  summarize(total = sum(n))

#Mutate table to calculate then append fractional word frequency in the text.
sortedwords.basic.fraction <- sortedwords.basic %>% as_tibble() %>% mutate(
  basic.fraction = (n*100)/total_words.basic$total
)
```

```{r}
sortedwords.writing <- base_data_tokenized.writing %>%
  filter(!word %in% stop_words$word,
         !word %in% str_remove_all(stop_words$word, "'"),
         str_detect(word, "[a-z]")) %>%
  count(word, word, sort = TRUE) %>%
  ungroup()

#Calculate the number of words in total in the subset
total_words.writing <- sortedwords.writing %>% 
  summarize(total = sum(n))

#Mutate table to calculate then append fractional word frequency in the text.
sortedwords.writing.fraction <- sortedwords.writing %>% as_tibble() %>% mutate(
  writing.fraction = (n*100)/total_words.writing$total
)

sortedwords.technical <- base_data_tokenized.technical %>%
  filter(!word %in% stop_words$word,
         !word %in% str_remove_all(stop_words$word, "'"),
         str_detect(word, "[a-z]")) %>%
  count(word, word, sort = TRUE) %>%
  ungroup()

#Calculate the number of words in total in the subset
total_words.technical <- sortedwords.technical %>% 
  summarize(total = sum(n))

#Mutate table to calculate then append fractional word frequency in the text.
sortedwords.technical.fraction <- sortedwords.technical %>% as_tibble() %>% mutate(
  technical.fraction = (n*100)/total_words.technical$total
)

sortedwords.logic <- base_data_tokenized.logic %>%
  filter(!word %in% stop_words$word,
         !word %in% str_remove_all(stop_words$word, "'"),
         str_detect(word, "[a-z]")) %>%
  count(word, word, sort = TRUE) %>%
  ungroup()

#Calculate the number of words in total in the subset
total_words.logic <- sortedwords.logic %>% 
  summarize(total = sum(n))

#Mutate table to calculate then append fractional word frequency in the text.
sortedwords.logic.fraction <- sortedwords.logic %>% as_tibble() %>% mutate(
  logic.fraction = (n*100)/total_words.logic$total
)
```

Next block joins the individual data tables for each subset in a larger file using column named 'word'. Columns get confusing names so they are renamed. **WARNING**: if a word is MISSING from **any subset** the data for that word is pruned from the full dataset. I am losing the unique words this way.

**Need to figure out how to join without pruning.**

```{r}
total <- merge(sortedwords.comments.all.fraction,sortedwords.technical.fraction,by="word") 
total2 <- merge(total,sortedwords.writing.fraction,by="word")
total3 <- merge(total2,sortedwords.logic.fraction,by="word")
#rename lines that were possibly confusing
names(total3)[2] <- "all.count"
names(total3)[4] <- "technical.count"
names(total3)[6] <- "writing.count"
names(total3)[8] <- "logic.count"

#Mutate table to calculate then append fractional word frequency in the text.
total4 <- total3 %>% as_tibble() %>% mutate(
  writing.wt = (writing.count *100)/all.count
)

total5 <- total4 %>% as_tibble() %>% mutate(
  technical.wt = (technical.count *100)/all.count
)

total6 <- total5 %>% as_tibble() %>% mutate(
  logic.wt = (logic.count *100)/all.count
)

```

The following six graphs compare frequencies for ~770 words that appear in ALL THREE of the sub-categories.

Graphs 1-3 compare frequencies in individual groups against word frequency overall. Generally, the frequency of a term in the whole set of comments is similar to frequency in an individual group (X vs Y positions similar.) However, some terms associated with Logic subgroup clearly fall below the trendline Graph 3, suggesting they are less frequent in that sub-category than overall.

```{r}
ggplot(total6, aes(`all.fraction`,`writing.fraction`)) +
  geom_jitter(alpha = 0.1, size = 2.5, width = 0.25, height = 0.25) +
  geom_text(aes(label = word), check_overlap = TRUE, vjust = 1.5) +
  scale_x_log10(labels = percent_format()) +
  scale_y_log10(labels = percent_format()) +
  geom_abline(color = "red")

ggplot(total6, aes(`all.fraction`,`technical.fraction`)) +
  geom_jitter(alpha = 0.1, size = 2.5, width = 0.25, height = 0.25) +
  geom_text(aes(label = word), check_overlap = TRUE, vjust = 1.5) +
  scale_x_log10(labels = percent_format()) +
  scale_y_log10(labels = percent_format()) +
  geom_abline(color = "red")

ggplot(total6, aes(`all.fraction`,`logic.fraction`)) +
  geom_jitter(alpha = 0.1, size = 2.5, width = 0.25, height = 0.25) +
  geom_text(aes(label = word), check_overlap = TRUE, vjust = 1.5) +
  scale_x_log10(labels = percent_format()) +
  scale_y_log10(labels = percent_format()) +
  geom_abline(color = "red")
```

Graphs 4-6 are pair-wise comparisons of frequencies for sub-categories. Unlike the comparisons to overall frequency, most words are NOT clustered along the trendline, suggesting there are greater differences in word usage frequencies.

```{r}
ggplot(total6, aes(`technical.fraction`,`writing.fraction`)) +
  geom_jitter(alpha = 0.1, size = 2.5, width = 0.25, height = 0.25) +
  geom_text(aes(label = word), check_overlap = TRUE, vjust = 1.5) +
  scale_x_log10(labels = percent_format()) +
  scale_y_log10(labels = percent_format()) +
  geom_abline(color = "red")

ggplot(total6, aes(`logic.fraction`,`writing.fraction`)) +
  geom_jitter(alpha = 0.1, size = 2.5, width = 0.25, height = 0.25) +
  geom_text(aes(label = word), check_overlap = TRUE, vjust = 1.5) +
  scale_x_log10(labels = percent_format()) +
  scale_y_log10(labels = percent_format()) +
  geom_abline(color = "red")

ggplot(total6, aes(`logic.fraction`,`technical.fraction`)) +
  geom_jitter(alpha = 0.1, size = 2.5, width = 0.25, height = 0.25) +
  geom_text(aes(label = word), check_overlap = TRUE, vjust = 1.5) +
  scale_x_log10(labels = percent_format()) +
  scale_y_log10(labels = percent_format()) +
  geom_abline(color = "red")
```

>>>>

---
title: "Data Import and Initial Exploration"
author: "Dan Johnson"
date: "12/17/2018"
output: html_document
bibliography: library.bib
biblio-style: apalike
---
```{r, child="_setup.Rmd"}
```
#Overview
This page has three sections. **Part 1** describes the target dataset for this project. **Part 2** outlines the data pre-checks and import process. **Part 3** describes post-import validation developed for the target dataset, and how selected elements within the dataset are recoded for analysis.

Initial visual exploration of the dataset is described here: https://adanieljohnson.github.io/default_website/initial_exploration.html

These linked pages summarize work to date using categorical Naive Bayes as the text classification method. They provided a baseline from which I am evaluating other approaches.

*  Baseline Method: https://adanieljohnson.github.io/default_website/Naive_bayesV2b.html
*  Randomization Method: https://adanieljohnson.github.io/default_website/Naive_bayesV2r.html
*  In/Outgroup Binning: https://adanieljohnson.github.io/default_website/Naive_bayesV2yn.html
*  In/Outgroup Binning: https://adanieljohnson.github.io/default_website/Naive_bayes_Smry.html

Other pages provide the theoretical background for the selected methods. Each deeper analysis method is described in detail on its own separate page. These are listed on the **[LINK] Project Outline** page.

@@

#Part 1. Initial Dataset
The development dataset is an Excel table of 10,808 rows. One column has "TA comments" extracted from student lab reports in Spring 2018. Each TA comment is stored as a separate row of the table, and is a mixed alphanumeric string of one or more words, numbers, and punctuation. Other columns record unique report, student, and TA IDs; grade assigned to the report; other standardized information about the original report from which the comment was extracted; and the hand-coded subject and structure of each comment.

Data were anonymized in Excel prior to conversion to CSV format. Names, email addresses, and other confidential information was replaced with unique randomized alphanumeric strings **[LINK]**.

@@

#Part 2. Data Pre-Checks
Prior to import, review the data table **in Excel**; check that:

1.  Student and TA names have been recoded using pre-assigned anonymous IDs.
2.  Vocabulary terms used in coding columns matches criteria in "codebook_for_comments.rmd"
3.  Data table headers match those listed below. The same list is in the file "code_column_names.rmd"

Column Number|Column Name|Accepted Values or Format
------------|----------------------|-----------------------
1|unique.record|Sp18.00001 (semester and year, then order in dataset)
2|report.id|R_0fanqHFhaY2yLT7 or similar
3|sort|remnant of coding
4|report.title (empty)|string
5|student|Std_drVZ2x1W (format is Std\_nnn)
6|course|113, 114, or 214
7|ta |TA_bc8EVX09 (format is TA\_nnn)
8|lab|113: allocation, betta; 114: manduca, frog; 214: enzymes (old) photosynthesis, physarum, cyanophage (new)
9|tag|first, second
10|type.TA|submission, revision
11|grade.TA|A, B, C, F
12|grading.time|time in minutes
13|Rank|remnant of coding
14|hypothesis.ok|Yes, No
15|data.ok|Yes, No
16|citation.ok|Yes, No
17|interpretation.ok|Yes, No
18|organization.ok|Yes, No
19|techflaws.ok|Yes, No
20|writing.ok|Yes, No
21|comments.incorporated|Yes, Somewhat, No
22|ta.comment|text string
23|code.subject|values in Ref. Tables
24|code.structure|values in Ref. Tables
25|code.locus|values in Ref. Tables
26|code.scope|values in Ref. Tables
27|code.tone|values in Ref. Tables
28|code.notes|values in Ref. Tables

@@

##Initial Library and Data Import
```{r Libraries_1}
# Call in tidyverse, other required libraries.
library(tidyverse)
library(tidytext)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
```

```{r}
#This block reads in full CSV into a starting dataframe named "base_data".
#Subsequent analyses are performed on extracted subsets of "base_data".
base_data <- read_csv(file='data/coded_full_comments_dataset_Spring18anonVERB.csv')
```

If dataset imports correctly, there will be a dataframe of ~11,000 rows and 28 columns named as listed in the **Data Pre-Checks** table above.

#Part 3. Post-Import Checks
Given some data points may be entered incorrectly by instructors or students, these post-import data checks help ensure data have been properly coded and columns contain valid entries only. (This process can be automated using the ```janitor``` data cleaning package.)

@@

#"Lab" should only have entries that match the allowed limited keywords in codebook.
lab<-unique(base_data$lab)

#It is impractical to validate all randomized identifiers. Instead look at the NUMBER of unique anonymous student IDs. Quantity should be within 10% of (but not more than) the combined enrollment of the 3 target courses for the semester.
student<-unique(base_data$student)

@@

#Check that the list of "code.structure" terms extracted from the dataset and displayed here matches allowed terms in codebook.
structure<-unique(base_data$code.structure)
```
####\

##Post-Import Processing
Comment labels used in the project codebook contain punctuation and spacing that complicates subsequent analysis. Rather than modify the project codebook or external processes, it is simpler to recode the problematic labels after import. For example, the following code extracts a subset of rows from the full dataset, reduces those to 3 columns, and renames the **comment.subject** column as **subject**. Four problematic code labels in the subject column are replaced with simpler terms.

```{r}
#Read in TA comments from CSV file.
base_data <- read_csv(file='data/coded_full_comments_dataset_Spring18anon.csv')

#Select rows representing the sub-groups to compare. 
comments_subset <- filter(base_data,code.subject=="1. Basic Criteria"|code.subject=="2. Writing Quality"|code.subject=="3. Technical and Scientific"|code.subject=="4. Logic and Thinking")

#Reduce larger dataframe to 3 required columns of data (subject, structure, and text of comment), and put columns in order needed.
comments_raw <- comments_subset %>% select(23,24,22)

#Rename the columns.
names(comments_raw)[1] <- "subject"
names(comments_raw)[2] <- "structure"
names(comments_raw)[3] <- "text"

#Simplify coding terms used in "subject" column.
comments_raw[,1] <- ifelse(comments_raw[,1] == "1. Basic Criteria","1_basic", ifelse(comments_raw[,1] == "2. Writing Quality","2_writing", ifelse(comments_raw[,1] == "3. Technical and Scientific","3_technical", ifelse(comments_raw[,1] == "4. Logic and Thinking","4_logic",99))))

#Change "subject" element from character to a factor for analysis.
comments_raw$subject <- factor(comments_raw$subject)
str(comments_raw$subject)
table(comments_raw$subject)
```

Most in-depth analyses will have a similar extraction and post-processing code block. In practice, it has proven easier to modify the post-processing code than to maintain multiple separate copies of the original CSV dataset for the different analyses.

### 2019-02-05 index.Rmd
Update

## Background
I am interested in understanding how students develop technical writing skills in biology, and more specifically, how we can apply data science methods to improve that process.

My current NSF-sponsored project asks if and how scripted instruction, automated support, and holistic feedback accelerates student development as writers. One of our specific aims is to see if changing the structure and focus of instructors' comments affects how students' writing develops over time. To do this, we need to extract instructor comments from student reports, classify those comments, and correlate the types of comments TAs make with student performance.

@@

* __Subject__: did the comment focus mainly on Basic criteria, Technical issues, Writing quality, Logic, or Other issues
* __Structure__: did the comment Point Out an error without providing other help, provide simple Copy Correction, provide declarative General or Specific Information only, or provide Holistic coaching that fosters student thinking and long-term improvement?

@@

#### \   
## Specific Project Goals
TA comments and meta-data for each student report can be extracted and compiled into a CSV datafile using existing Unix scripts. However routinely classifying TA comments by hand is impractical; 10-12,000 comments need to be scored EACH SEMESTER. An automated classification method is needed both for speed and consistency.

My specific project goals were/are to create an R-based text analysis workflow that:

1. **Converts TA comments and metadata into a de-identified "tidy" dataframe for R.** This includes converting the master CSV data set to an anonymized data frame with all names and confidential metadata re-keyed to unique random identifiers.
2. **Uses text features to assign TA comments to the categories defined in our existing codebook**. The goal is to be able to assign instructor comments into our pre-defined Subject and Structure codes with 90% accuracy or greater.

@@

>>>>

## Background
I am interested in understanding how students develop technical writing skills in biology, and more specifically, how to apply data science methods to improve that process.

My current NSF-sponsored project asks if and how scripted instruction, automated support, and holistic feedback accelerates student development as writers. One specific aim is to see if changing the structure and focus of instructors' comments affects how students' writing develops over time. To measure this we extract instructor comments from student reports, classify those comments, and correlate the types of comments TAs make with student performance.

@@

* __Subject__: did the comment focus mainly on Basic criteria, Technical issues, Writing quality, Logic and thinking, or Other issues
* __Structure__: did the comment Point to an error without providing other help, provide simple Copy Correction, provide declarative General or Specific Information only, or provide Holistic coaching that fosters student thinking and long-term improvement?

@@

(Embed a copy of that table here.)

@@

## Specific Project Goals
Unix scripts can extract and compile TA comments and meta-data for each student report into a CSV datafile. However routinely classifying TA comments by hand is impractical; 10-12,000 comments need to be scored EACH SEMESTER. An automated classification method is needed both for speed and consistency.

My specific project goals were/are to:

1. **Create an R-based workflow that converts TA comments and metadata from student reports into a de-identified "tidy" dataframe for R.** This includes converting the master CSV data set to an anonymized data frame with all names and confidential metadata re-keyed to unique random identifiers.
2. **Use text features to assign TA comments to the hand-coded categories defined in our existing codebook**. The goal is to be able to assign instructor comments into our pre-defined Subject and Structure codes with 90% accuracy or greater. 
3. **Assemble these methods into a semi-automated text classifier**. The final classifier model will be able to generate a two-axis contingency table similar to the one created using hand-coded data. It will be applied to newer datasets to determine if and how TA comments change over semesters.
